\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-06 misc/JSS article/final/article.Rnw
Improve langauge.

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions---such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots---functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. Lastly, it includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

This exemplifies the power of these tools to create, query, and combine nested data in very general ways.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes, \fct{[ir]reg\_elem} to recursively extract the (non-)atomic elements, and \fct{rapply2d} to apply functions to nested lists of data objects.
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The function \fct{varying} can be used to examine which of these variables are time-varying.
%
<<varying>>=
varying(wlddev, ~ iso3c)
@
%
A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated.
%
<<pdec>>=
LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
@
%
The \fct{qsu} (quick-summary) function provides an efficient method to compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the overall mean back to the within transformation to preserve the scale of the data.
%
<<qsu_1>>=
qsu(LIFEEXi)
@
%
The decomposition above implies more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups and with sampling weights. \fct{qsu} also has a data frame method, and by default computes simple statistics:\footnote{The \code{pid} argument to \fct{qsu} can also be used to pass identifiers for panel-decomposition, e.g., \code{pid = ~ iso3c}. With indexed data, it is automatically set to the first column in the index (\code{effect = 1}).}
%
<<qsu_2>>=
qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
@
%
% The output shows that the variation in life expectancy is significantly larger for non-OECD countries. For the latter the between- and within-country variation is approximately equal.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX $\sim$ OECD, $\sim$ iso3c, $\sim$ POP)}, without prior indexation.}
For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description. It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very useful with mixed-type data.
%
<<descr>>=
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 |> descr(by = income + LIFEEX ~ OECD)
@

%
 It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'click-through' the variables in a large dataset while spreading the computational burden. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline

>>>>

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. It also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

This exemplifies the power of these tools to create, query, and combine nested data in very general ways.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes, \fct{[ir]reg\_elem} to recursively extract the (non-)atomic elements, and \fct{rapply2d} to apply functions to nested lists of data objects.
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}. The function \fct{varying} can be used to examine which of these variables are time-varying.
%
<<varying>>=
varying(wlddev, ~ iso3c)
@
%
A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated:
%
<<pdec>>=
LIFEEXi <- reindex(wlddev$LIFEEX, index = wlddev$iso3c)
all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
@
%
The \fct{qsu} (quick-summary) function provides an efficient method to compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the overall mean back to the within transformation to preserve the scale of the data.
%
<<qsu_1>>=
qsu(LIFEEXi)
@
%
The decomposition above implies more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups and with sampling weights. \fct{qsu} also has a data frame method, and by default computes simple statistics:\footnote{The \code{pid} argument to \fct{qsu} can also be used to pass identifiers for panel-decomposition, e.g., \code{pid = ~ iso3c}. With indexed data, it is automatically set to the first column in the index (\code{effect = 1}).}
%
<<qsu_2>>=
qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
@
%
% The output shows that the variation in life expectancy is significantly larger for non-OECD countries. For the latter the between- and within-country variation is approximately equal.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX $\sim$ OECD, $\sim$ iso3c, $\sim$ POP)}, without prior indexation.}
For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description. It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very useful with mixed-type data.
%
<<descr>>=
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 |> descr(by = income + LIFEEX ~ OECD)
@

%
 It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis. \newline
